Identification of rice genes which respond to saline stress from co-expression networks analysis

### 2020-09-11 camila/Report/article_CR.tex
paper draft

@@

We propose a methodology to identify stress responsive genes to salt conditions in rice. The methodology is based on Weighted Gene Co-expression Network Analysis (WGCNA). This is considered an effective and accurate bioinformatic method, using co-expression networks, that has been widely applied in identifying target genes for disease and cancer fields~\cite{tian2018identifying}. We follow the WGCNA workflow but with a new approach in the module detection step. Our modules are detected using the Hierarchical Link Clustering (HLC) technique~\cite{ahn2010link} that allows the recognition of overlapping communities, which may have more biological meaning given the overlapping regulatory domains of systems that generate co-expression~\cite{gaiteri2014beyond}. We conduct a systematic study with a large set of rice data using the proposed methodology. RNA-seq data was accessed through GEO database~\cite{GEOAcces90:online} (Accession number GSE98455), corresponding to $57845$ gene expression profiles of shoot tissues measured for both control and salt condition in $92$ accessions of the Rice Diversity Panel 1. As the analysis result, 6 modules are detected as relevant in the response to salt stress in rice: 3 modules of 3 genes each one associated with shoot K content, 2 modules of 3 genes associated with  shoot biomass, and 1 module of 4 genes associated with root biomass. These genes may act as potential targets for the improvement of salinity tolerance in rice cultivars. From those 19 genes, all but 3 genes (associated with $K$ content), were also identified as deferentially expressed ($|LFC| > 2$) for at least one of the 92 accessions, suggesting that those genes are strong candidates as stress responsive genes. Only 2 of the 16 diferentially expressed genes, both from the module related with shoot biomass, are named and have an associated protein product: Spermidine hydroxycinnamoyltransferase 2 (SHT2) and Lipoxygenase. In other words, further studies are needed to elucidate the detailed biological function of the remaining 14 genes that have not been named so far,  which may have a potential relevance in stress responsive mechanisms to salt conditions in rice.\\

@@

@@

This methodology uses as input data RNA-seq read counts, representing gene expression levels. More precisely, $n$ gene expression profiles of an organism, measured for $m$ different genotypes under control and treatment conditions, and $r$ biological replicates. This data can be represented as a matrix $D=[d_{ij}]$, where $d_{ij} \in \mathbb{N}_0$ for all $i = 1,2,..n$ and $j=1,2,..,2mr$. To discover key genes and their interaction for treatment-tolerance related phenotypes, the methodology also requires a set of $p$ phenotypic traits, measured for the $m$ genotypes. The phenotypic data can be seen as a matrix $P=[p_{jk}]$, where $p_{jk} \in \mathbb{R}$ for all $k = 1,2,...,p$.

@@

The RNA-seq data cannot be directly interpreted, therefore a normalization process has to be done to deal with the various biases that affect quantification results. In order to correct library size and RNA composition bias, the suggested normalization technique is DESeq2 \cite{love2014moderated}. Then, from the normalized data, average the biological replicates of each accession and remove genes exhibiting low variance or low expression. Next, separate control and salinity stress treatment data into two matrices $C=[c_{ij}]_{n \times m}$ and $T=[t_{ij}]_{n \times m}$, respectively, where $c_{ij}$ and $t_{ij}$ represent normalized expression level of the gene $i$ in the accession $j$.\\

Next step is to compile the control and treatment data by measuring the changes in expression levels in terms of logarithmic ratios. Matrix $L=[\ell_{ij}]_{n \times p}$, known as the Log Fold Change matrix, is computed by setting $\ell_{ij}=\log_2 (t_{ij}/c_{ij})$. Finally, filter $L$ matrix by removing rows (genes) with low variance in the differential expression patterns.

@@

The Log fold change matrix $L$ is used to build the co-expression network following the first steps of the WGCNA methodology. First, measure the level of concordance between gene differential expression profiles across samples. Use the absolute value of the Pearson correlation coefficient as the similarity measure between genes and store the values in the similarity matrix $S=[s_{ij}]_{n\times n}$.\\

Then, transform $S$ into an adjacency matrix $A=[a_{ij}]_{n\times n}$ where $a_{ij} = (s_{ij})^\beta $ encodes the connection strength between each pair of genes. In other words, the elements of the adjacency matrix are the similarity values up to the power $\beta > 1$ so the degree distribution will fit a scale-free network. This kind of networks contain many nodes with very few connections and a small number of hubs with high connections. In a strict scale-free network the logarithm of $P(k)$ (the probability of a node to have degree $k$) is approximately inversely proportional to the logarithm of $k$ (the degree of a node). So the parameter $\beta$ is chosen as the smallest value of $\beta$ such that the $R^2$ of the linear regression between $log_{10}(p(k))$ and $log_{10}(k)$ is close to $1$ (e.g. $R^2 > 0.85$). \\

Finally, prepare matrix $A$ to apply the clustering algorithm, transforming it into an unweighted network $\hat{A}$. To determine the Pearson Correlation Coefficient (PCC) cutoff for finding biologically relevant co-expressed modules, we use the approach described by~\cite{aoki2007approaches} based on density of the network combined with decreasing number of nodes and edges with higher PCC values.\\

The matrix $\hat{A}$ can be seen as an undirected graph $G=(V,E)$ where ${V=\{v_1,v_2,\ldots,v_n\}}$ is a set of vertices or nodes and ${E=\{e_1,e_2,\ldots,e_p\}}$ is a set of edges or links that connect the vertices. In this genomic context, the graph $G$ represents a co-expression network, where each node corresponds to a gene and a pair of genes is connected if they show similar differential expression patterns.

@@

Next step in the methodology is to study the co-expression network structure and dynamics identifying communities also called modules. The idea is to cluster genes with similar differential expression change patterns. Membership in these modules may overlap in biological contexts, where modules may be related to specific molecular, cellular or tissue functions and the biological components (i.e. genes) are involved in multiple functions. Thus, unlike WGCNA, the co-expression network $G$ is used to detect overlapping rather than non-overlapping communities, using the Hierarchical Link Clustering (HLC) algorithm proposed in~\cite{ahn2010link}.\\

HLC approach reinvent communities as groups of links rather than nodes, each node inherits all memberships of its links and can thus belong to multiple, overlapping communities. The algorithm maps links to nodes and connects them if a pair of links shares a node. They compute the similarity between links using the Jaccard index for unweighted networks. With this similarity, they use single-linkage hierarchical clustering to build a dendrogram where each leaf is a link from the original network and branches represent link communities. Finally, the most relevant communities are established at the maximal partition density, a function  based on link density inside communities.

@@

To identify the most relevant gene groups (modules), associated with the phenotypic response to a specific treatment in an organism, we use a LASSO based approach. Each module can be represented by a eigengene, which is defined as the first principal component of such module. A eigengene can be thought of as an average differential expression profile for each community and is computed from the Log Fold Change matrix rows corresponding to the genes belonging to a specific module. \\

These profiles are then associated with each phenotypic trait using the least absolute shrinkage and selection operator (LASSO). LASSO combines a regression model with a procedure of contraction of some parameters towards zero, imposing a restriction or a penalty on the regression coefficients. This technique allows selection of the most relevant variables. In our context, the eigenges act as regressor variables and each phenotypic trait its used as an outcome variable. \\

@@

You also can perform a functional category enrichment looking for the Gene Ontology (GO) annotations from databases like QuikGO~\cite{binns2009quickgo}. This annotations can provide evidence of biological implications of the target genes in the treatment-tolerance mechanisms.\\

Those named genes (with GO annotations), can be used to performe another relevant analysis reviewing their reported protein-protein interaction networks using the STRING database~\cite{szklarczyk2016string}. The interactions include direct (physical) and indirect (functional) associations; they stem from computational prediction, from knowledge transfer between organisms, and from interactions aggregated from other (primary) databases. This information elucidates how the selected genes are involved in functional pathways that can be related with the treatment of interest.

\section{Case study}
RNA-seq data was accessed through GEO database \cite{GEOAcces90:online} (Accession number GSE98455), corresponding to $n=57845$ gene expression profiles of shoot tissues measured for both control and salt condition in $m=92$ accessions of the Rice Diversity Panel 1, with two biological repetitions ($r=2$). \\

Genes exhibiting low variance (the ratio of upper quantile to lower quantile smaller than 1.5) or low expression (more than 80\% samples with values smaller than 10) were removed. The final network contains $8928$ genes of the initial $57845$.\\

Figure~\ref{fig:beta} shows the degree distribution of the similarity matrix (left) and the degree distribution of the adjacency matrix (right) which is the degree distribution of a scale-free forced network with $R^2 = 0.8$ corresponding to $\beta = 3$.

@@

The current network represented by the adjacency matrix $A$, corresponds to a complete and weighted network of $8928$ genes (nodes) and $39850128$ edges. For computational reasons, this network was transformed into an unweighted one $\hat{A}$, keeping only the connections above the cutoff value of $0.2$. The resulting unweighted network has a total of $5810$ nodes and $16875145$ edges. After applying the HLC algorithm, a total of $4131$ genes were distributed in $5143$ overlapping modules of $3$ or more genes. Figure~\ref{fig:overlap} shows a histogram of the overlapping percentage of these genes, measured as the proportion of modules to which each gene belongs.

%overlapping percentage
\begin{figure}[h]
  \centering
    \includegraphics[clip,width=0.4\textwidth]{Figures/artificial_modules.png}
  \caption{Overlapping percentage}
  \label{fig:overlap}
\end{figure}

As shown in Figure \ref{fig:pdata}, there are significant differences in the values of these phenotypic traits between both stress and control conditions. This supports the idea that these variables represent tolerance-associated traits in rice under salt stress.

@@

We use 3 phenotypic traits: shoot $K^+$ content, root biomass and shoot biomass. These were measured for the 92 genotypes studied, under  control and salt stress conditions. These data can be found in the supplementary information of~\cite{campbell2017allelic}.\\

Finally, 6 modules were detected as relevant in the response to salt stress in rice: 3 modules of 3 genes each one associated with shoot K content, 2 modules of 3 genes associated with  shoot biomass, and 1 module of 4 genes associated with root biomass.
From those 19 genes, all but 3 genes (associated with $K$ content), were also identified as deferentially expressed ($|LFC| > 2$) for at least one of the 92 accessions. Only 2 of the 16 diferentially expressed genes, both from the module related with shoot biomass, are named and have an associated protein product: Spermidine hydroxycinnamoyltransferase 2 (SHT2) and Lipoxygenase.

@@

@@

\newpage
\section*{Appendix}
\subsection{A: Variable selection with LASSO}
LASSO (Least Absolute Shrinkage Selector Operator) is a regularized linear regression technique, a method that combines a regression model with a procedure of contraction of some parameters towards zero and selection of variables, imposing a restriction or a penalty on the regression coefficients. Very useful in problems where the number of variables (genes) $ n $ is much greater than the number of samples $ p $ ($ n \gg p $). Lasso solves the least squares problem with restriction on the $ L_1$-norm of the coefficient vector:

\begin{equation}
\min \left\lbrace\sum_{i=1}^{p}{\left( y_i-\sum_{j=1}^n{\beta_j x_{ij}}\right)^2} \right\rbrace , \textrm{sujeto a} \sum_{j=1}^n\abs{\beta_j}\leq s
\end{equation}

Since the $\lambda$ value determines the degree of penalty, the accuracy of the model depends on its choice. Cross-validation is often used to select the regularization parameter, choosing the one that minimizes the mean-squared error. With that selected $\lambda$ value, the model is adjusted again, this time using all the observations.\\

In the module gene selection context, the outcome variable correspond to the phenotypic trait, whereas the predictors are the modules, detected by the hierarchical clustering, represented by the first principal component of the module. After running the Lasso regression will select the most significant modules associated with phenotypic data.

>>>>

\usepackage{subcaption} % subfigures

@@

We propose a methodology to identify stress responsive genes to salt conditions in rice. The methodology is based on Weighted Gene Co-expression Network Analysis (WGCNA). This is considered an effective and accurate bioinformatic method, using co-expression networks, that has been widely applied in identifying target genes for disease and cancer fields~\cite{tian2018identifying}. We follow the WGCNA workflow but with a new approach in the module detection step. Our modules are detected using the Hierarchical Link Clustering (HLC) technique~\cite{ahn2010link} that allows the recognition of overlapping communities, which may have more biological meaning given the overlapping regulatory domains of systems that generate co-expression~\cite{gaiteri2014beyond}. We conduct a systematic study with a large set of rice data using the proposed methodology. RNA-seq data was accessed through GEO database~\cite{GEOAcces90:online} (Accession number GSE98455), corresponding to $57845$ gene expression profiles of shoot tissues measured for both control and salt condition in $92$ accessions of the Rice Diversity Panel 1. As the analysis result, 6 modules are detected as relevant in the response to salt stress in rice: 3 modules of 3 genes each one associated with shoot K content, 2 modules of 3 genes associated with  shoot biomass, and 1 module of 4 genes associated with root biomass. These genes may act as potential targets for the improvement of salinity tolerance in rice cultivars. From those 19 genes, all but 3 genes (associated with $K$ content), were also identified as deferentially expressed (with a log fold change value greater than $2$) for at least one of the 92 accessions, suggesting that those genes are strong candidates as stress responsive genes. Only 2 of the 16 diferentially expressed genes, both from the module related with shoot biomass, are named and have an associated protein product: Spermidine hydroxycinnamoyltransferase 2 (SHT2) and Lipoxygenase. In other words, further studies are needed to elucidate the detailed biological function of the remaining 14 genes that have not been named so far,  which may have a potential relevance in stress responsive mechanisms to salt conditions in rice.\\

@@

\section{Preliminaries}

\subsection{Co-expression network}
A network is an undirected graph $G=(V,E)$ where ${V=\{v_1,v_2,\ldots,v_{n}\}}$ is a set of vertices or nodes and ${E=\{e_1,e_2,\ldots,e_q\}}$ is a set of edges or links that connect the vertices. In a gene co-expression network each node corresponds to a gene and a pair of genes is connected if they show similar differential expression patterns. These co-expression networks are of biological interest since the co-expressed genes are usually controlled by the same transcriptional regulatory pathway, functionally related or members of the same pathway or metabolic complex.\\

When the network is simple and unweighted, it can be represented by an adjacency matrix $A \in \{0,1\}^{n \times n}$. This adjacency matrix is symmetric, with a positive one in the positions $(v_i,v_j)$ and $(v_j,v_i)$ whenever there is an edge connecting vertices $v_i$ and $v_j$, and zeros elsewhere.\\

\subsection{Hierarchical Link Clustering}

Hierarchical Link Clustering (HLC) algorithm was proposed by Ahn et al.~\cite{ahn2010link}. HLC approach reinvent communities as groups of links rather than nodes, each node inherits all memberships of its links and can thus belong to multiple, overlapping communities. The algorithm maps links to nodes and connects them if a pair of links shares a node. The similarity between two links $e_{ik}$ and $e_{jk}$ is computed using the Jaccard index 
\begin{equation*}
\label{eq:jaccard}
S(e_{ik},e_{jk}) = \frac{\vert n_+(i) \cap n_+(j) \vert}{\vert n_+(i) \cup n_+(j) \vert},
\end{equation*}
where $n_+(i)$ denotes the set of node $i$ and its neighbors.\\

With this similarity, the method uses single-linkage hierarchical clustering to build a dendrogram where each leaf is a link from the original network and branches represent link communities. Hierarchical clustering methods repeatedly merge groups until all elements are members of a single cluster. To find meaningful communities it is crucial to know where to partition the dendrogram. Thus, the most relevant communities are established at the maximal partition density $D$, a function  based on link density inside communities that measures the quality of a link partition. The partition density has a single global maximum along the dendrogram in almost all cases, because the value is just the average density at the top of the dendrogram (a single giant community with every link and node) and it is very small at the bottom of the dendrogram (most communities consists of a single link). $D = 1$ when every community is a fully connected clique and $D = 0$ when each community is a tree. If a link community is less dense than a tree (when the community subgraph has disconnected components), then that community will give a negative contribution to $D$. Computing $D$ at each level of the link dendrogram allows to pick the best level to cut (although meaningful structure exists above and below that threshold). This way, the final output corresponds to node clusters, where each node can participate in multiple communities.\\

\subsection{Least Absolute Shrinkage Selector Operator}
Least Absolute Shrinkage Selector Operator (LASSO) is a regularized linear regression technique, a method that combines a regression model with a procedure of contraction of some parameters towards zero and selection of variables, imposing a restriction or a penalty on the regression coefficients. Very useful in problems where the number of variables (genes) $ n $ is much greater than the number of samples $ m $ ($ n \gg m $). Lasso solves the least squares problem with restriction on the $ L_1$-norm of the coefficient vector:

\begin{equation}
\min \left\lbrace\sum_{i=1}^{m}{\left( y_i-\sum_{j=1}^n{\beta_j x_{ij}}\right)^2} \right\rbrace , \textrm{subject to} \sum_{j=1}^n\abs{\beta_j}\leq s
\end{equation}

Since the $\lambda$ value determines the degree of penalty, the accuracy of the model depends on its choice. Cross-validation is often used to select the regularization parameter, choosing the one that minimizes the mean-squared error. With that selected $\lambda$ value, the model is adjusted again, using all the observations.\\

@@

This methodology uses as input data RNA-seq read counts, representing gene expression levels. More precisely, $n_0$ gene expression profiles of an organism, measured for $m$ different genotypes under control and treatment conditions, and $r$ biological replicates. 
This raw data can be represented as a matrix $D_0 \in {\mathbb{N}_0}^{n_0 \times 2mr}$. 
In order to discover key genes and their interaction with phenotypes related to treatment tolerance, the methodology also requires a set of $p$ phenotypic traits, measured for the $m$ genotypes. The phenotypic data can be seen as a matrix $P \in \mathbb{R}^{2m \times p}$ containing two phenotypic values per genotype, one under control condition and the second under treatment condition.\\

@@

The RNA-seq data cannot be directly interpreted, therefore a normalization process has to be done to deal with the various biases that affect quantification results. In order to correct library size and RNA composition bias, the suggested normalization technique is DESeq2 \cite{love2014moderated}. From the normalized data $D_1 \in \mathbb{R}^{n_0 \times 2mr}$, average the biological replicates of each genotype, so that the data matrix becomes $D_2 \in \mathbb{R}^{n_0 \times 2m}$. Next, remove genes exhibiting low variance or low expression, getting a new number of $n_1 \leq n_0$ genes. Separate control and treatment data into two matrices $C\in \mathbb{R}^{n_1 \times m}$ and $T\in \mathbb{R}^{n_1 \times m}$, respectively. Each matrix entry, $c_{ij}$ for control and $t_{ij}$ for treatment, represent the normalized expression level of gene $i$ in accession $j$.\\

Now, compile the control and treatment data by measuring the changes in expression levels in terms of logarithmic ratios. Matrix $L_0 \in \mathbb{R}^{n_1 \times m}$, known as the Log Fold Change matrix, is computed by setting $\ell_{ij}=\log_2 (t_{ij}/c_{ij})$. Repeat this process on the phenotypic data separating the control and treatment data into matrices $P_c$ and $P_t$ of dimensions $m \times p$. Calculate the corresponding logarithmic ratios, obtaining the matrix $P_\ell \in \mathbb{R}^{m \times p}$. Finally, filter $L_0$ matrix by removing rows (genes) with low variance in the differential expression patterns, getting a new matrix $L_1$ of dimensions $n_2 \times m$, with $n_2 \leq n_1$.\\

@@

The Log fold change matrix $L_1$ is used to build the co-expression network following the first steps of the WGCNA methodology. First, measure the level of concordance between gene differential expression profiles across samples. Use the absolute value of the Pearson correlation coefficient as the similarity measure between genes and store the values in the similarity matrix $S\in \mathbb{R_{+}}^{n_2 \times n_2}$.\\

Then, transform $S$ into an adjacency matrix $A \in \mathbb{R_+}^{n_2\times n_2}$ where each entry $a_{ij} = (s_{ij})^\beta $ encodes the connection strength between each pair of genes. In other words, the elements of the adjacency matrix are the similarity values up to the power $\beta > 1$ so the degree distribution will fit a scale-free network. This kind of networks contain many nodes with very few connections and a small number of hubs with high connections. In a strict scale-free network the logarithm of $P(k)$ (the probability of a node to have degree $k$) is approximately inversely proportional to the logarithm of $k$ (the degree of a node). So the parameter $\beta$ is chosen as the smallest value of $\beta$ such that the $R^2$ of the linear regression between $log_{10}(p(k))$ and $log_{10}(k)$ is close to $1$ (e.g. $R^2 > 0.85$). \\

Finally, prepare matrix $A$ to apply the clustering algorithm, transforming it into an unweighted network $\hat{A} \in \{0,1\}^{n_2 \times n_2}$. To determine the Pearson Correlation Coefficient (PCC) cutoff for finding biologically relevant co-expressed modules, we use the approach described by~\cite{aoki2007approaches} based on density of the network combined with decreasing number of nodes and edges with higher PCC values. 
Above the cutoff, the entries of matrix $A$ become $1$ and when the value is below the threshold it becomes zero.\\

@@

Next step in the methodology is to study the co-expression network structure and dynamics identifying communities also called modules. The idea is to cluster genes with similar differential expression change patterns. Membership in these modules may overlap in biological contexts, where modules may be related to specific molecular, cellular or tissue functions and the biological components (i.e. genes) are involved in multiple functions. Thus, unlike WGCNA, the adjacency matrix $\hat{A}$, from the co-expression network $G$, is used to detect overlapping rather than non-overlapping communities, using the Hierarchical Link Clustering (HLC) algorithm proposed in~\cite{ahn2010link}.\\

%HLC approach reinvent communities as groups of links rather than nodes, each node inherits all memberships of its links and can thus belong to multiple overlapping communities. The algorithm maps links to nodes and connects them if a pair of links shares a node. The similarity between links is computed using the Jaccard index. With this similarity, they use single-linkage hierarchical clustering to build a dendrogram where each leaf is a link from the original network and branches represent link communities. Finally, the most relevant communities are established at the maximal partition density, a function  based on link density inside communities.\\

The HLC algorithm organizes the $n_2$ genes of matrix $\hat{A}$ into $c$ communities, where each gene can belong to one, multiple, or no community. This information is represented as an affiliation matrix $F \in \{0,1\}^{n_2 \times c}$, where $f_{iu} = 1$ if node $i$ is member of cluster $u$.\\

@@

To identify the most relevant gene groups (modules), associated with the phenotypic response to a specific treatment in an organism, we use a LASSO based approach. Each module can be represented by a eigengene, which is defined as the first principal component of such module. A eigengene can be thought of as an average differential expression profile for each community and is computed from the Log Fold Change Matrix $L_1$ and the affiliation matrix $F$. Given a module $u$, use the affiliation matrix to identify the genes belonging to $u$ and then select the corresponding rows of the matrix $L_1$ to compute the first principal component of $u$. Each principal component would be a column of the matrix $M \in \mathbb{R}^{m \times c}$.\\

% the Log Fold Change matrix rows corresponding to the genes belonging to a specific module. \\

These profiles are then associated with each phenotypic trait using the least absolute shrinkage and selection operator (LASSO). LASSO combines a regression model with a procedure of contraction of some parameters towards zero, imposing a restriction or a penalty on the regression coefficients. This technique allows selection of the most relevant variables. In our context, the eigenges (the columns of $M$) act as regressor variables and each phenotypic trait (each column of $P_\ell$) is used as an outcome variable. \\

@@

You also can perform a functional category enrichment searching for the Gene Ontology (GO) annotations from databases like QuikGO~\cite{binns2009quickgo}. This annotations can provide evidence of biological implications of the target genes in the treatment-tolerance mechanisms.\\

Those named genes (with GO annotations), can be used to perform another relevant analysis reviewing their reported protein-protein interaction networks using the STRING database~\cite{szklarczyk2016string}. The interactions include direct (physical) and indirect (functional) associations; they stem from computational prediction, from knowledge transfer between organisms, and from interactions aggregated from other (primary) databases. This information elucidates how the selected genes are involved in functional pathways that can be related with the treatment of interest.\\

\section{Case study}
RNA-seq data was accessed through GEO database \cite{GEOAcces90:online} (Accession number GSE98455), corresponding to $n_0=57845$ gene expression profiles of shoot tissues measured for both control and salt condition in $m=92$ accessions of the Rice Diversity Panel 1, with $r=2$ biological replicates. On the other hand, we use $p=3 $ phenotypic traits: shoot $K^+$ content, root biomass and shoot biomass. This traits were measured for the same $92$ genotypes, under  control and salt stress conditions, and can be found in the supplementary information of~\cite{campbell2017allelic}.\\

\subsection{Data pre-processing}
The DESeq normalization was applied to the raw data and the biological replicates were averaged. Genes exhibiting low variance were identified as those with ratio of upper quantile to lower quantile smaller than $1.5$, and were removed. Genes with low expression were also removed, corresponding to those having more than $80\%$ samples with values smaller than $10$. After this filtering process, $n_1 = 8928$ genes were kept.\\

From the compiled matrix $L_0$ we remove genes whose difference between upper quantile and lower quantile is greater than $0.25$. So the resulting matrix $L_1$ contains the log ratios of $n_2 = 8928$ genes. We also compute the logarithmic ratios of the phenotipic data, for the $92$ accessions and the $3$ traits.\\

\subsection{Co-expression network construction}
We used the Log Fold Change matrix $L_1$ to compute the corresponding similarity matrix. 
We found that for this network, the smallest integer such that the $R^2 \geq 0.8 $ is $\beta=3$. Figure~\ref{fig:beta} shows the degree distribution of the similarity matrix (left) and the degree distribution of the adjacency matrix (right) which is the degree distribution of a scale-free forced network with $R^2 = 0.8$ corresponding to $\beta = 3$.

@@

The resulting adjacency matrix $A$ represents a complete graph $G=(V,E)$, with $|V| = 8928$ genes (nodes) and $|E| = 39850128$ edges. We choose $0.2$ as cutoff value, keeping only the connections above this threshold and removing the isolated nodes, so the new adjacency matrix $\hat{A}$ has $n_3 = 5810$ genes and accounts for $16875145$ edges.\\

%The current network represented by the adjacency matrix $A$, corresponds to a complete and weighted network of $8928$ genes (nodes) and $39850128$ edges. For computational reasons, this network was transformed into an unweighted one $\hat{A}$, keeping only the connections above the cutoff value of $0.2$. The resulting unweighted network has a total of $5810$ nodes and $16875145$ edges.

\subsection{Co-expression module identification}
After applying the HLC algorithm, a total of $4131$ genes were distributed in $c = 5143$ overlapping modules of $3$ or more genes. Figure~\ref{fig:overlap} shows a histogram of the overlapping percentage of these genes, measured as the proportion of modules to which each gene belongs.

%overlapping percentage
\begin{figure}[h]
  \centering
    \includegraphics[clip,width=0.5\textwidth]{Figures/artificial_modules.png}
  \caption{Overlapping percentage}
  \label{fig:overlap}
\end{figure}

\subsection{Modules association to phenotypic trits}
The phenotypic traits under study are shoot $K^+$ content, root biomass and shoot biomass. Figure \ref{fig:pdata} shows that there are significant differences in the values of these phenotypic traits between stress and control conditions. This supports the idea that these variables represent tolerance-associated traits in rice under salt stress.

@@

Using the affiliation matrix $F$ derived from HLC output and the Log Fold Change matrix $L_1$, we build the matrix $M$ computing the egengene for each of the $c = 5143$ modules.
We apply the LASSO technique using each of the phenotypic traits as the outcome variable, one at a time. As shown in Figure~\ref{fig:cross-val} we perform cross-validation for each phenotypical trait, in order to select the corresponding regularization parameter $\lambda$ that minimizes the mean-squared error.\\

\begin{figure}[h]
\centering
   \begin{subfigure}{0.3\linewidth} \centering
     \includegraphics[width=1\textwidth]{Figures/lambda_kshoot_2.png}
     \caption{$K$ shoot, $\lambda = 0.013$}\label{fig:lambda_kshoot}
   \end{subfigure}
   \begin{subfigure}{0.3\linewidth} \centering
     \includegraphics[width=1\textwidth]{Figures/lambda_BMshoot_2.png}
     \caption{Shoot biomass, $\lambda = 0.025$}\label{fig:lambda_BMshoot}
   \end{subfigure}
   \begin{subfigure}{0.3\linewidth} \centering
     \includegraphics[width=1\textwidth]{Figures/lambda_BMroot.png}
     \caption{Root biomass, $\lambda = 0.035$}\label{fig:lambda_BMroot}
   \end{subfigure}
\caption{Cross-validation of the LASSO regularization parameter $\lambda$, for each phenotypic trait.} \label{fig:cross-val}
\end{figure}

We finally adjust the three models using all observations and the corresponding $\lambda$ values. As result, 6 modules were detected as relevant in the response to salt stress in rice: 3 modules of 3 genes each one associated with shoot $K$ content, 2 modules of 3 genes associated with  shoot biomass, and 1 module of 4 genes associated with root biomass.\\

\subsection{Genes enrichment}
From the $19$ genes selected by LASSO, all but $3$ genes (associated with $K$ content), were also identified as deferentially expressed ($|\ell_{ij}| \geq 2$) for at least one of the $92$ accessions. This suggest that those genes are strong candidates as stress responsive genes to salt conditions in rice.\\

According to the Quickgo database, only $2$ of the $16$ diferentially expressed genes, both from the module related with shoot biomass, are named and have an associated protein product: Spermidine hydroxycinnamoyltransferase 2 (SHT2) and Lipoxygenase. Figure~\ref{fig:3d} shows their corresponding 3D protein structure.

% Figure 3d protein strucures
\begin{figure}[h]
\centering
   \begin{subfigure}{0.49\linewidth} \centering
     \includegraphics[width=0.5\textwidth]{Figures/structure_LOC_Os12g27254.png}
     \caption{Spermidine hydroxycinnamoyltransferase 2}\label{fig:structure_SHT2}
   \end{subfigure}
   \begin{subfigure}{0.49\linewidth} \centering
     \includegraphics[width=0.5\textwidth]{Figures/structure_LOC_Os12g37260.png}
     \caption{Lipoxygenase}\label{fig:structure_Lipoxygenase}
   \end{subfigure}
\caption{3D protein structure of named genes selected by LASSO.} \label{fig:3d}
\end{figure}

On the one hand, SHT2 contributes to the natural variation of spermidine-based phenolamides in rice cultivars. On the other hand plant lipoxygenase may be involved in a number of diverse aspects of plant physiology including growth and development, pest resistance, and senescence or responses to wounding.This protein is involved in the pathway oxylipin biosynthesis, which is part of Lipid metabolism.\\

Previous studies (\cite{gupta2013plant, hou2015persimmon, mittova2002salt, peng2019novel, roychoudhury2011amelioration}) provide evidence of biological implications of sperimidine and lipoxygenase in tolerance to salt stress in other plants or even in rice cultivars. However, further studies are needed to elucidate the detailed biological function of the remaining 14 genes that have not been named so far,  which may have a potential relevance in stress responsive mechanisms to salt conditions in rice.

%We also use the STRING database to visualize the reported interaction networks of the 2 named genes (see Figure X).

%Figure STRING networks

@@

The methodology described in this manuscript provide an aproach to discover key genes responding to a specific treatment in an organism, linking transcriptomic with phenotypic data, and identifying overlaping gene modules. Similar approximations have been proposed~(\cite{langfelder2008wgcna, du2019network}), but they only focus on disjoint (non-overlapping) communities. However, it is well known that communities in real networks are overlapping. \\

The methodology was applied in a case study with rice under salt stress. The results show a group of 14 genes where two of them are related to the response to saline stress according to previous studies, validating the ability of the method to detect this kind of key genes.\\

An advantage of the proposed methodology is its modularity. As future work, other module detection and selection techniques could be used, instead HLC and LASSO respectively. The combinations of these techniques would allow finding target genes for future biological studies that evaluate their potential as genes that respond to salt stress in rice. Furthermore, this study can be extended to other stresses and even to other crops.\\

@@
